---
title: "Class11"
author: 'Camryn McCann (PID: A15437387)'
date: "11/2/2021"
output: pdf_document
---

#Introduction to the RCSB Protein Data Bank (PDB)

Downloaded the following CSV dile from the PDB site.

```{r}
db <- read.csv("Data Export Summary.csv", row.names = 1)
head(db)
```

> **Q1: What percentage of structures in the PDB are solved by X-Ray and Electron Microscopy.**

```{r}
round(sum(db$X.ray)/sum(db$Total)*100,2)
```

```{r}
round(sum(db$EM)/sum(db$Total)*100,2)
```

> **Q2: What proportion of structures in the PDB are protein?**

```{r}
round(db$Total[1]/sum(db$Total)*100,2)
```

> **Q3: Type HIV in the PDB website search box on the home page and determine how many HIV-1 protease structures are in the current PDB?**

????

#Visualizing the HIV-1 protease structure

> **Q4: Water molecules normally have 3 atoms. Why do we see just one atom per water molecule in this structure?**

These water molecules only have one atom, which is oxygen, because the 2 hydrogen atoms present are too small to see.

> **Q5: There is a conserved water molecule in the binding site. Can you identify this water molecule? What residue number does this water molecule have (see note below)?**

![](vmdscene.tga)

The water molecule is conserved at residue 308.

#Introduction to Bio3D in R

```{r}
library(bio3d)
```

```{r}
pdb <- read.pdb("1hsg.pdb")
pdb
```


#Comparative structure analysis of Adenylate Kinase