---
title: 'Class 10: Halloween Candy'
author: 'Camryn McCann (PID: A15437387)'
date: "10/28/2021"
output: html_document
---

#DATA

First, lets import and look at our data

```{r}
candy_file <- "candy-data.csv"
candy = read.csv(candy_file, row.names=1)
head(candy)
```
According to 538 the columns in the dataset include:

chocolate: *Does it contain chocolate?*
fruity: *Is it fruit flavored?*
caramel: *Is there caramel in the candy?*
peanutyalmondy: *Does it contain peanuts, peanut butter or almonds?*
nougat: *Does it contain nougat?*
crispedricewafer: *Does it contain crisped rice, wafers, or a cookie component?*
hard: *Is it a hard candy?*
bar: *Is it a candy bar?*
pluribus: *Is it one of many candies in a bag or box?*
sugarpercent: *The percentile of sugar it falls under within the data set.*
pricepercent: *The unit price percentile compared to the rest of the set.*
winpercent: *The overall win percentage according to 269,000 matchups (more on this in a moment).*

>**Q1. How many different candy types are in this dataset?** 

```{r}
nrow(candy)
```

>**Q2. How many fruity candy types are in the dataset?**

```{r}
sum(candy[,2])
#both these ways can tell us answer for the fruity column
sum(candy$fruity)

```

Let's look at winpercent! 

>**Q3. What is your favorite candy in the dataset and what is it’s winpercent value?**

My favorite is 3 Musketeers... let's see...

```{r}
candy["3 Musketeers", ]$winpercent
```

>**Q4. What is the winpercent value for “Kit Kat”?**

```{r}
candy["Kit Kat", ]$winpercent
```

>**Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?**

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

*There is a helpful skim function in the skimr package that can help us overview the data set*

```{r}
library(skimr)
skim(candy)
```
Using the results of the skim function, we can answer the following questions. 

>**Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?**

winpercent is on a different scale than the majority of the other columsn in the data set

>**Q7. What do you think a zero and one represent for the candy$chocolate column?** 

0 represents that the candy does not have chocolate
1 represents that the candy does have chocolate

#Now lets look at this visually by making a plot.

>**Q8. Plot a histogram of winpercent values**

```{r}
hist(candy$winpercent)
#we can also do this on ggplot
library(ggplot2)
ggplot(candy, aes(winpercent)) + geom_histogram(bins= 10,col= "black", fill= "orange")
```

>**Q9. Is the distribution of winpercent values symmetrical?**

No, the distribution is not symmetrical.

>**Q10. Is the center of the distribution above or below 50%?**

Below 50%

>**Q11. On average is chocolate candy higher or lower ranked than fruit candy?** 

First let's look at all the chocolate

```{r}
choc <- as.logical(candy$chocolate)
chocolate <- candy[choc,]$winpercent
```

Now let's look at the fruity candy

```{r}
fruit <- as.logical(candy$fruity)
fruity <- candy[fruit,]$winpercent
```

We can look at the average of each 

```{r}
mean(chocolate)
mean(fruity)
```

On average, chocolate candy is ranked higher than fruity candy. 

>**Q12. Is this difference statistically significant?**

```{r}
t.test(chocolate, fruity)
```

Because the p-value is < 0.05, there is a significant difference between the winpercent of chocolate and fruity candy. 

# Looking at overall candy rankings

>**Q13. What are the five least liked candy types in this set?**

```{r}
head(candy[order(candy$winpercent),], n=5)
```
The five least liked candies are in order from least liked to most are Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters. 

>**Q14. What are the top 5 all time favorite candy types out of this set?**

```{r}
ord <- order(candy$winpercent, decreasing=TRUE)
head(candy[ord,], n=5)
```

>**Q15. Make a first barplot of candy ranking based on winpercent values.**

```{r}
library(ggplot2)
ggplot(candy) + 
  aes(winpercent, rownames(candy), winpercent) +
  geom_col()
```

>**Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?**

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col()
```

Now let's add some useful color aesthetics and plot again.

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

Using this graph we can answer the following: 

>**Q17. What is the worst ranked chocolate candy?**

Sixlets 

>**What is the best ranked fruity candy?**

Starburts


# Let's look at candy pricepoint 

```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

>**Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?** 


>**Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?**

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```
The top 5 most expensive candies in the data set are Nik L Nip, Nestle Smarties, Ring pop, Hershey's Krackel, and Hershey's Milk Chocolate. Of these, the least popular is Nik L Nip with a win percent of only 22%. 


#Exploring Correlation Structure

```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```


>**Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?** 

The chocolate and fruity variables are anti-correlated. 

>**Q23. Similarly, what two variables are most positively correlated?** 

Chocolate and Bar are most positively correlated to each other.

#Lastly, lets do some PCA

```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)
```



